### 2010-03-31 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: write All and all using the first definition of Desc.
Desc@ICFP: fix elaboration of constructors in Mu

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throws away information when we
crucially need it. In the realm of data-types, things become even
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a sea of $\Sigma$s, our data-types have little
chance of survival. 
\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of codes is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational
machinery. \note{Is that a good description?}

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from a
Haskell-like $\data$ definition.

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its codes, $\DSigma{\!}{\!}$,
$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
theory. Hence, in an implementation of this type theory, we need to
extended it with these constructors and their typing rules. Although
it makes no conceptual difference, the code of description should
rather be read as a specification. We promise the existence of such
objects, satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

So far, we have used our universe of descriptions to build
functors. \note{ Strictly positive types anyone? } To illustrate its
usage, we have implemented the pattern functors of natural numbers,
lists, and binary trees. However, the class of functors expressible
with descriptions enjoys another property: they all admit a
fix-point. Indeed, descriptions are interpreted into \emph{polynomial
  functors}~\cite{who?}\note{who is the relevant source for polynomial
  functors?}, as they are only composed of $\Sigma$-types, the unit
type, and variables.

@@

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to address
the second issue. Namely, the developer would rather be presented a
name than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write data-type constructors is simply a
matter of type propagation. We therefore extend the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:

@@

Hence, in the high-level expression language, the developer writes a
familiar type constructor, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.

>>>>

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throw away information when we
crucially need it. In the realm of data-types, things become even more
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a sea of $\Sigma$s, our data-types have little
chance of survival. 
\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of code is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational
machinery. \note{Is that a good description?}

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from an
Haskell-like $\data$ definition.

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{\!}{\!}$,
$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
theory. Hence, in an implementation of this type theory, we need to
extended it with these constructors and their typing rules. Although
it makes no conceptual difference, the code of description should
rather be read as a specification. We promise the existence of such
objects, satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

So far, we have used our universe of descriptions to build
functors. \note{ Strictly positive types anyone? } To illustrate its
usage, we have implemented the pattern functors of natural numbers,
lists, and binary trees. However, the class of functors expressible
with descriptions enjoy another property: they all admit a
fix-point. Indeed, descriptions are interpreted into \emph{polynomial
  functors}~\cite{who?}\note{who is the relevant source for polynomial
  functors?}, as they are only composed of $\Sigma$-types, the unit
type, and variables.

@@

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to tamper
the second issue. Namely, the developer would rather be presented a
name rather than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write data-type constructor is simply a
matter of type propagation. We therefore extends the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:

@@

Hence, in the high-level expression language, the developer writes a
familiar type constructors, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.
